The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Natural or tidal?

### 2025-09-24 thesis/abstract.tex
half-done discussion

The satellite galaxies of the Milky Way (MW), with the exception of the
Magellanic Clouds, are all dwarf spheroidals (dSphs)---gas-free,
non-rotating stellar systems whose gravity is dominated by dark matter.
The light profiles of dSphs are usually well-approximated by an
exponential law, with a sharp outer cutoff. Yet systems like the
Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
stars far from the main body, out to \(\sim10\) effective radii,
indicative of outer deviations from an exponential law. The origin of
these extended profiles is not clear, and may reflect either the
influence of Galactic tides or, instead, an innate feature. In this
thesis, we review the existence of extended light profiles in Scl and
UMi, validating the Bayesian membership catalogue of
\citet{jensen+2024}. To evaluate if tides indeed can explain these
density features, we conduct idealized N-body simulation of both
galaxies in the tidal field of the Milky Way. We find that neither dwarf
has been subjected to tidal forces strong enough to create the observed
extended light profiles from initially exponential profiles. One
complication is that Sculptor's orbit is strongly influenced by the
presence of the Large Magellanic Cloud (LMC). Because the LMC's mass is
poorly constrained, Sculptor's long term orbital history is highly
uncertain. Despite this uncertainty, our models still suggest that the
combined tides of the LMC and MW are unable to explain Scl's outer
profile. We conclude that the extended light profiles of Scl and UMi are
innate, likely reflecting their assembly histories. Mergers, accretion
events, or complex, multi-component star formation episodes are possible
explanations for the observations ed extended light profiles.

>>>>

The satellite galaxies of the Milky Way (MW), except the Magellanic
Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating
stellar systems dynamically dominated by dark matter. The light profiles
of dSphs are usually well-approximated by an exponential law, with a
steep outer cutoff. Yet systems like the Sculptor (Scl) and Ursa Minor
(UMi) dSphs have distant member stars, out to \(\sim10\) effective radii
from the centre, indicative of outer deviations from an exponential law.
The origin of these extended profiles is unclear, and may reflect either
the influence of Galactic tides or an innate feature. In this thesis, we
review the existence of extended light profiles in Scl and UMi,
validating the Bayesian membership catalogue of \citet{jensen+2024}. To
evaluate if tides indeed can explain these density features, we conduct
idealized N-body simulation of both galaxies in the tidal field of the
Milky Way. We find that neither dwarf has been subjected to tidal forces
strong enough to create the observed extended light profiles from
initially exponential profiles. One complication is that Sculptor's
orbit is strongly influenced by the presence of the Large Magellanic
Cloud (LMC). Our models still suggest that the combined tides of the LMC
and MW are unable to explain Scl's outer profile. We conclude that the
extended light profiles of Scl and UMi are innate, likely reflecting
their assembly histories. Mergers, accretion events, or episodic star
formation are possible explanations for the observations of extended
light profiles.

### 2025-09-24 thesis/data.tex
half-done discussion

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second component
is included only when the preferred amplitude of the second component is
non-zero. The \emph{proper-motion likelihood} quantifies the agreement
of a star's motion with the dwarf galaxy's systemic motion, accounting
for observational uncertainties. The \emph{CMD likelihood} measures the
consistency of a star's \emph{Gaia} \(G\) magnitude and
\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
galaxy. For the background likelihoods, the spatial likelihood is a
uniform distribution over the field, and the background CMD and PM
likelihoods are constructed empirically from the stars in an annulus far
from the satellite. Each likelihood is normalized as a probability
density over the respective parameter space.

@@

We analyze stellar distributions in the tangent plane, considering as
well the projected shape of a galaxy. The tangent plane coordinates
\(\xi\) and \(\eta\) are offsets in RA and declination as measured on
the plane tangent relative to the galaxy centre. To account for the
elliptical shape of the galaxy, we use \(R_{\rm ell}\), which we define
as the circularized elliptical radius, \begin{equation}{
R_{\rm ell}^2 = a\,b\,\left(\frac{{\xi'}^2}{a^2} + \frac{{\eta'}^2}{b^2} \right),
}\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
coordinate rotated to align with a dwarf galaxy's major and minor axis,
and \(a\) and \(b\) are the semi-major and semi-minor axis of the
galaxy.

@@

Based on the fiducial sample's distribution in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
Ursa Minor stands out because of its higher ellipticity. Sculptor and
Ursa Minor show a population of members past \(6R_h\) in radius, but
Fornax does not, even though Fornax has many more stars. Fornax also has
a bluer CMD, indicative of more recent star formation. Otherwise, there
are no major morphological differences between these galaxies.

One limitation of the J+24 method is the assumption of a specific
density profile for the dwarf galaxy (a one- or two-component
exponential), which may impact the membership probability of distant
stars. However, there are no clear extensions or overdensities in the
``CMD+PM'' selected sample, which does not include a spatial likelihood.
If even fainter, more extended features exist, they are not clearly
detectable with \emph{Gaia}.

@@

We derive density profiles by binning member stars in constant width
bins in \(\log R_{\rm ell}\) of 0.05 dex. We ignore bins interior or
exterior to the first empty bin in either direction. We use symmetric
Poisson uncertainties as ``error bars'' in the density estimate at each
bin.

@@

While the density profiles of all samples agree in the inner regions,
they begin to deviate outside \(6 R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the CMD+PM selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix \ref{sec:density_extra}.

Near UMi, there is a small (\(R_h\sim 0.5'\)), likely unassociated,
ultrafaint star cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a
relative position of \((\xi, \eta) \approx(-42, -15)\) arcminutes,
corresponding to an elliptical radius of 37 arcminutes. This cluster
does not have a bright RGB, so has few stars brighter than a magnitude
\(G=22\). The cluster has little effect on the elliptically-averaged
density profile (see location on Fig.~\ref{fig:scl_observed_profiles}).

>>>>

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second component
is included only when the preferred amplitude of the second component is
non-zero. The \emph{proper-motion likelihood} quantifies the agreement
of a star's motion with the dwarf galaxy's systemic motion, accounting
for observational uncertainties. The \emph{CMD likelihood} measures the
consistency of a star's \emph{Gaia} \(G\) magnitude and
\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
galaxy. For the background likelihoods, the spatial likelihood is a
uniform distribution over the field, and the background CMD and PM
likelihoods are constructed empirically from the stars in an annulus far
from the satellite. Each likelihood is normalized as a probability
density over the respective parameter space. The appendix discusses the
details of quality cuts and the likelihood calculation.

@@

We analyze stellar distributions in the tangent plane and consider the
projected shape of a galaxy. The tangent plane coordinates \(\xi\) and
\(\eta\) are offsets in right ascension (RA) and declination as measured
on the plane tangent relative to the galaxy centre. To account for the
elliptical shape of the galaxy, we use \(R_{\rm ell}\), which we define
as the circularized elliptical radius, \begin{equation}{
R_{\rm ell}^2 = a\,b\,\left(\frac{{\xi'}^2}{a^2} + \frac{{\eta'}^2}{b^2} \right),
}\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
coordinate rotated to align with a dwarf galaxy's major and minor axis,
and \(a\) and \(b\) are the semi-major and semi-minor axis of the
galaxy.

@@

Based on the fiducial sample's distribution in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
Ursa Minor stands out because of its higher ellipticity. While Sculptor
and Ursa Minor show an extended population of members past \(6R_h\) in
radius Fornax does not despite having many more stars. Fornax also has a
bluer CMD, indicative of more recent star formation. Otherwise, there
are no major morphological differences between these galaxies.

One limitation of the J+24 method is the assumption of a specific
density profile for the dwarf galaxy (a one- or two-component
exponential), which may impact the membership probability of distant
stars. However, there are no clear extensions or overdensities in the
``CMD+PM'' selected sample, which does not include a spatial likelihood.
If even fainter, more extended features exist, they are not obviously
detectable with \emph{Gaia}.

@@

We derive density profiles by binning member stars in constant width
bins in \(\log R_{\rm ell}\) of 0.05 dex. We ignore bins interior or
exterior to the first empty bin in either direction. We use symmetric
Poisson uncertainties as error bars in the density estimate at each bin.

@@

While the density profiles of all samples agree in the inner regions,
they begin to deviate outside \(6 R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the ``CMD+PM'' selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix \ref{sec:density_extra}.

Near UMi, there is a small (\(R_h\sim 0.5'\)), likely unassociated,
ultrafaint star cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a
relative position of \((\xi, \eta) \approx(-42, -15)\) arcminutes,
corresponding to an elliptical radius of 37 arcminutes. Muñoz 1 does not
have a bright RGB, so has few stars above the \emph{Gaia} magnitude
limit. The cluster has little effect on the elliptically-averaged
density profile (see location on Fig.~\ref{fig:scl_observed_profiles}).

### 2025-09-24 thesis/discussion.tex
julio edits at durham
half-done discussion

\subsection{Long term orbital history}\label{long-term-orbital-history}

@@

@@

\section{Comparison to other work}\label{comparison-to-other-work}

@@

\begin{itemize}
\tightlist
\item
  \citet{battaglia+2008}
\item
  \citet{iorio+2019}
\item
  \citet{amorisco+zavala+deboer2014}
\item
  \citet{battaglia+2008}
\item
  \citet{breddels+2013}
\item
  \citet{breddels+helmi2013}
\item
  \citet{richardson+fairbairn2014}
\item
  \citet{SFW2017}
\item
  \citet{innanen+papp1979}
\item
  \citet{wilkinson+2002}
\item
  \citet{yang+2025}: chemical evolution in Scl.
\item
  \citet{skuladottir+2024}; \citet{skuladottir+2021};
  \citet{lee+jeon+bromm2024}, pop III high res spectro?
\item
  \citet{wang+2024a} hydrosim of dwarf galaxies like sculptor
\item
  \citet{tang+2023}: apogee modeling of Scl.
\item
  \citet{pandey+west2022} chem evo OMEGA and isotopes.
\item
  \citet{an+koposov2022}: distance gradients?
\item
  \citet{kawata+2006}, 2pop origins?
\item
  \citet{grcevich+putman2009}, \citet{carignan+1998} ambiguous HI clouds
  near Scl ?
\item
  \citet{agnello+evans2012}, claim that Sculptor cannot hold two
  populations in NFW halo?
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \citet{sestito+2023a}
\item
  \citet{westfall+2006} wide degree survey for extended structure.
\item
  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
  \citet{arroyo-polonio+2024}
\item
  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
\item
  \citet{coleman+dacosta+bland-hawthorn2005}
\item
  \citet{DQ1994}
\item
  \citet{WMO2009}
\item
  \citet{IH1995},
\item
  \citet{munoz+2018}: Using Megacam to derive density profiles and
  structural properties of many dwarf spheroidal galaxies.
\item
  \citet{kirby+2009}
\item
  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
\item
  \citet{grebel1996}
\item
  \citet{barbosa+2025}: Using DECam to derive narrowband photometric
  metallicity gradient and search for metal poor stars in Scl.
\end{itemize}

Future ideas:

\begin{itemize}
\tightlist
\item
  \citet{evslin2016}: measuring ellipticities of halos w TMT.
\end{itemize}

@@

\section{Are Sculptor and Ursa Minor
typical?}\label{are-sculptor-and-ursa-minor-typical}

\subsection{The formation of Exponential
profiles}\label{the-formation-of-exponential-profiles}

As mentioned earlier, many dwarf galaxies appear to have exponential
stellar density profiles.

@@

Given that dwarf galaxies were noticed to be exponential like above,
\citet{faber+lin1983} proposed that dwarf spheroidals formed from disky
galaxies maintaining the typical exponential density profile

@@

In summary, while there is some theoretical reasons why exponential
spheroidal stellar profiles may form, the emperical sucess of an
exponential law still requires explanations. Deviations from this
exponential thus test wether this exponential law is indeed
near-universal or if other ingredients in the formation and history of
dwarf galaxies may be responsible for changes from the empirical
expectation.

\subsection{Alternatives to exponential
profiles}\label{alternatives-to-exponential-profiles}

\begin{itemize}
\tightlist
\item
  elmegreen+hunter2006: Discuss the formation of double exponentials in
  disk galaxies. While a simpler idealized / Semi-analytic model of star
  formation, show that assuming exponentail gas and KS-relation with SF
  threshhold, that reduction of turbulance (driven by a number of
  processes) in outer disk leads to a steeper decline in SFR as crossing
  the critical SF threshold becomes less likely.
\end{itemize}

@@

Are dwarf galaxies indeed expected to be one-component exponential-like
density profiles? The suggestion of exponential density profiles dates
back to \citet{faber+lin1983}. Expand\ldots{}

@@

\begin{itemize}
\tightlist
\item
  \citet{revaz+jablonka2018}
\item
  \citet{tau+vivas+martinez-vazquez2024} (RRL stars )
\item
  \citet{roderick+2016} Boötes I
\item
  \citet{mcconnachie+penarrubia+navarro2007} (velocity space \& multiple
  populations).
\item
  \citet{cicuendez+battaglia2018}
\end{itemize}

\subsection{Multi-epoch star formation and
feedback}\label{multi-epoch-star-formation-and-feedback}

@@

\begin{itemize}
\tightlist
\item
  \citet{maxwell+2012}
\item
  \citet{wright+2019}
\item
  \citet{azartash-namin+2024}
\end{itemize}

\subsubsection{Multi-component density
profiles}\label{multi-component-density-profiles}

\begin{itemize}
\tightlist
\item
  \citet{benitez-llambay+2016}
\item
  \citet{mercado+2021}: formation of metallicity gradients from SFH
\item
  \citet{revaz+jablonka2018} natural formation of multi-component and
  gradients from cosmological simulations of dwarf galaxies. Originate
  from dynamical heating of metal rich population and outside in star
  formation?
\item
  \citet{el-badry+2016}. Feedback drives fluctuations in size + radial
  migration =\textgreater{} population gradients.
\item
\end{itemize}

\begin{itemize}
\tightlist
\item
  \citet{arroyo-polonio+2024}
\item
  \citet{pace+2020}
\item
  \citet{fabrizio+2016}, \citet{kordopatis+2016} (multi-component in
  Carina)
\item
  \citet{battaglia+2006}, \citet{amorisco+evans2012} for fornax
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \citet{deason+2014}
\item
  \citet{deason+2022}
\item
  \citet{ricotti+polisensky+cleland2022}
\item
  \citet{querci+2025}
\item
  \citet{amorisco+evans+vandeven2014}
\item
  \citet{tarumi+yoshida+frebel2021}: Shows cosmological simulations in
  context of Tuc II, merger of two galaxies in the early universe
  produces a more extended density profile and metallicity gradient.
\item
  \citet{lokas+2014}: And II from merger,
\item
  rotation from merger: \citet{cardona-barrero+2021}
\end{itemize}

@@

Another alternative is that the dwarf galaxies have been
``preprocessed'' by another dwarf/LMC like galaxy earlier on and this
has resulted in their unique properties

@@

>>>>

\subsection{Challenges to backwards-time
integration}\label{sec:scl_orbit_variability}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_lmc_orbits_mass_effect.png}}
\caption[Sculptor Orbits with LMC]{The long-term orbital history of
Sculptor is uncertain. In light, transparent lines, the orbits of
Sculptor in three different LMC/MW mass models from \citet{vasiliev2024}
are shown for random samples of the observed properties. The LMC orbits
are in solid, thick lines of the corresponding colour. The L2M11 has a
lighter LMC mass, and the L3M10 model has a lighter MW mass than our
fiducial L3M11 LMC model.}\label{fig:scl_lmc_orbits_mass}
\end{figure}

Our LMC model only simulates recent tidal effects---the long-term orbit
of Scl is essentially unknown. Fig.~\ref{fig:scl_lmc_orbits_effect}
illustrates the orbits of Scl for three different LMC-MW models in
\citet{vasiliev2024}: L2M11 has a lighter LMC and L3M10 has a lighter MW
than the L3M11 potential we used. All orbits diverge significantly after
2 Gyr ago. In addition, other affects such as the lumpiness of the MW
halo, the growth of the MW halo, dynamical friction, and limitations of
this potential make long-term backwards time integration yet more
unreliable \citep[e.g.,][]{dsouza+bell2022}.

In Fig.~\ref{fig:scl_lmc_orbits_effect}, some orbits have a small
\(\lesssim 10\,\kpc\) pericentre with the Milky Way. In the Appendix, we
consider a model with a previous MW pericentre of 4kpc. While this
encounter should affect the inner stellar distribution of Sculptor based
on the Jacobi radius, the observed light profile remains unaffected.
Because this model experiences just one tidal shock, stronger tidal
streams form, but the tidal impact is not strong enough to reshape the
observed stellar distribution. We conclude that

As discussed in \citet{vasiliev2024}, if the LMC has had a past, second
pericentre with the Milky Way, there is a chance that UMi was previously
bound to the LMC. We find that this occurs about 15\% of the time for
this choice of potential. However, backwards integration may not
appropriately represent the true probability of a Magellanic origin,
since tidal stripping is not time reversible. Because of the possibility
of this past encounter, and the high uncertainties associated with
backwards integration, we note this as a possibility but defer to the
discussion for our interpretations.

@@

\section{Comparison to previous work}\label{comparison-to-previous-work}

For both Sculptor and Ursa minor, these galaxies have been studied
extensively in both a theoretical and observational context.

@@

\begin{itemize}
\tightlist
\item
  \citet{battaglia+2008}
\item
  \citet{iorio+2019}
\item
  \citet{amorisco+zavala+deboer2014}
\item
  \citet{breddels+2013}
\item
  \citet{breddels+helmi2013}
\item
  \citet{richardson+fairbairn2014}
\item
  \citet{SFW2017}
\item
  \citet{innanen+papp1979}
\item
  \citet{wilkinson+2002}
\item
  \citet{yang+2025}: chemical evolution in Scl.
\item
  \citet{skuladottir+2024}; \citet{skuladottir+2021};
  \citet{lee+jeon+bromm2024}, pop III high res spectro?
\item
  \citet{wang+2024a} hydrosim of dwarf galaxies like sculptor
\item
  \citet{tang+2023}: apogee modeling of Scl.
\item
  \citet{pandey+west2022} chem evo OMEGA and isotopes.
\item
  \citet{an+koposov2022}: distance gradients?
\item
  \citet{kawata+2006}, 2pop origins?
\item
  \citet{grcevich+putman2009}, \citet{carignan+1998} ambiguous HI clouds
  near Scl ?
\item
  \citet{agnello+evans2012}, claim that Sculptor cannot hold two
  populations in NFW halo?
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \citet{tolstoy+2004}: detection of two populations.
\item
  \citet{battaglia+2008a}
\item
  \citet{sestito+2023a}
\item
  \citet{westfall+2006} wide degree survey for extended structure.
\item
  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
  \citet{arroyo-polonio+2024}
\item
  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
\item
  \citet{coleman+dacosta+bland-hawthorn2005}
\item
  \citet{DQ1994}
\item
  \citet{WMO2009}
\item
  \citet{IH1995},
\item
  \citet{munoz+2018}: Using Megacam to derive density profiles and
  structural properties of many dwarf spheroidal galaxies.
\item
  \citet{kirby+2009}
\item
  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
\item
  \citet{grebel1996}
\item
  \citet{barbosa+2025}: Using DECam to derive narrowband photometric
  metallicity gradient and search for metal poor stars in Scl.
\end{itemize}

@@

\section{Should we expect exponential
profiles?}\label{should-we-expect-exponential-profiles}

As discussed in Section~\ref{sec:exponential_profiles}, the exponential
profile has long been used for dwarf galaxies.

@@

@@

elmegreen+hunter2006: Discuss the formation of double exponentials in
disk galaxies. While a simpler idealized / Semi-analytic model of star
formation, show that assuming exponentail gas and KS-relation with SF
threshhold, that reduction of turbulance (driven by a number of
processes) in outer disk leads to a steeper decline in SFR as crossing
the critical SF threshold becomes less likely.

@@

@@

\begin{itemize}
\tightlist
\item
  \citet{revaz+jablonka2018}
\item
  \citet{tau+vivas+martinez-vazquez2024} (RRL stars )
\item
  \citet{roderick+2016} Boötes I
\item
  \citet{mcconnachie+penarrubia+navarro2007} (velocity space \& multiple
  populations).
\item
  \citet{cicuendez+battaglia2018}
\item
  \citet{delpino+aparicio+hidalgo2015}: Fornax
\item
  \citet{delpino+2017}, \citet{delpino+2017a}
\end{itemize}

\subsection{Multi-component density profiles: clues to the
past?}\label{multi-component-density-profiles-clues-to-the-past}

\begin{itemize}
\item
  \citet{benitez-llambay+2016}, \citet{benitez-llambay+2015}
\item
  \citet{mercado+2021}: formation of metallicity gradients from SFH
\item
  \citet{revaz+jablonka2018} natural formation of multi-component and
  gradients from cosmological simulations of dwarf galaxies. Originate
  from dynamical heating of metal rich population and outside in star
  formation?
\item
  \citet{el-badry+2016}. Feedback drives fluctuations in size + radial
  migration =\textgreater{} population gradients.
\end{itemize}

\begin{itemize}
\tightlist
\item
  \citet{arroyo-polonio+2024}
\item
  \citet{pace+2020}
\item
  \citet{battaglia+2012}, \citet{fabrizio+2016}, \citet{kordopatis+2016}
  (multi-component in Carina)
\item
  \citet{battaglia+2006}, \citet{amorisco+evans2012} for fornax
\item
  \citet{battaglia+2011} sextans
\item
  \citet{ho+2012}: And II
\end{itemize}

\subsection{Multi-epoch star
formation}\label{multi-epoch-star-formation}

@@

\begin{itemize}
\item
  \citet{maxwell+2012}
\item
  \citet{wright+2019}
\item
  \citet{azartash-namin+2024}
\item
  \citet{dong+lin+murray2003}
\end{itemize}

\subsection{Dynamical heating of old
stars}\label{dynamical-heating-of-old-stars}

For a variety of reasons, older stars in a galaxy may be preferentially
``hotter'' (i.e., higher typical random velocities) than younger stars.

In the case of dwarf galaxies, a few mechanisms have been proposed which
may heat the stellar components: heating by strong feedback, heating by
dark sub-subhalos, and heating

Because dwarf galaxies have extended star formation histories, like the
stellar-feedback scattering method,

One compl

\emph{cluster dynamics} Could processess which happen in globular
clusters also occur in dsarf galaxies, such as mass segragation and the
evaporation of the smallest mass stars? This seems unlikely at first
glance because dwarf galaxies contain such

\citet{baumgardt+2022}.

@@

When galaxies merge, traces of the effects of the merger can remain for
long afterwards. For dwarf galaxies, mergers may appear as multiple
populations which may have different sizes, chemistry, and kinematics.

A number of studies has investigated mergers in dwarf galaxies, showing
that this is a feasible scenario for halo formation

\begin{itemize}
\tightlist
\item
  \citet{deason+2014}
\item
  \citet{deason+2022}
\item
  \citet{ricotti+polisensky+cleland2022}
\item
  \citet{querci+2025}
\item
  \citet{amorisco+evans+vandeven2014}
\item
  \citet{tarumi+yoshida+frebel2021}: Shows cosmological simulations in
  context of Tuc II, merger of two galaxies in the early universe
  produces a more extended density profile and metallicity gradient.
\item
  \citet{lokas+2014}: And II from merger,
\item
  rotation from merger: \citet{cardona-barrero+2021}
\item
  \citet{starkenburg+helmi+sales2016}
\item
  \citet{fouquet+2017}: And II via merger
\item
  \citet{genina+2019}: \apostle{} galaxies: Mergers with gas can heat
  old population and form new population subsequently. Additionally,
  ram-pressure stripping and interactions with cosmic filimants can
  remove outer gas and trigger inner star formation.
\end{itemize}

@@

Another alternative is that the dwarf galaxies have been
``preprocessed'' by another dwarf/LMC like galaxy earlier on and this
has resulted in their unique properties.

This could involve a strong tidal interaction very early on in the dwarf
galaxy.

@@

\section{Summary and outlook}\label{summary-and-outlook}

In this thesis, we have investigated tidal effects on the dwarf galaxies
Sculptor and Ursa Minor.

We first showed, using data from J+24, that each galaxy contains a
substantial, extended population when compared to an exponential density
profile. We show that this population is reasonably independent of the
details of the data selection or methodology, implying that this
``density excess'' is likely a real feature of each galaxy. Sculptor and
Ursa Minor are more extended in the outer regions than other nearby
dwarf classical, given current observations.

We then investigated if tides were a permissable explanation. By
modelling each galaxy based on cosmological initial conditions, we
showed that tides do not strongly affect either galaxy. However, the
presence of an LMC in the MW possibly complicates our conclusions, but
we find that the effect of the LMC means Scl could be on first infal and
has experienced a significantlly weaker tidal history, and Ursa Minor
may only have a more extended orbit. In all of our investigation, we
show that tides are very unlikely to affect the currently-observed
density profiles. A stream is possibly detectable, depending on the
initial stellar extent of each galaxy and how much deeper future
observations are able to reach.

As a result, we conclude that the observed extended density of each
galaxy is inconsistent with a tidal origin, and requires a different
explanation. In this (final) chapter, we have summarized possible other
explanations in the literature, including the effects of mergrs,
episodic star formation histories, stellar halos,

While we have discussed a number of different avenues through which
stellar halos form, these need not be exclusive.

As an example, consider +frenk+carlos+++: when two gas-rich dwarf
galaxies collide, the merger can introduce new star formation, possibly
forming a new, distint population of stars. The formation of stars may
drive a burst which heats up old stars. The merger may scatter old stars
into a stellar halo. And the merger may distribute the two populations
of the old galaxies into separate components. As a result, many
different processes can contribute to the formation of chemically and
dynamically distinct populations.

Is there hope for determining the nature of a stellar halo? Perhaps the
most promising tool will be chemistry. In particular, if we reach a
point where we can chemically distinguish the populations of different
galaxies, then chemistry provides an opportunity to test the merger
history of galaxies.

We propose a categorization of the (\LCDM{}) scenarios above

\begin{itemize}
\tightlist
\item
  Gas-poor mergers or accretion: more than one \emph{chemodynamically
  independent} population (e.g.~GSE in the MW halo)
\item
  Gas-rich merger: more than one \emph{chemodynamically independent}
  population plus an additional, younger population with possibly shared
  heritage.
\item
  Continuous dynamical heating (e.g.~subsubhalos) : chemically
  contiguous population with an old to young population gradient.
\item
  Strong stellar feedback: Associated bursty star formation history in
  chemical abundance space, but reasonably chemically-contiguous
  population.
\end{itemize}

In addition, the first two scenarios could show significant variation
among dwarfs (especially for (near) major-mergers), whereas the last two
should be relatively similar among similar dwarfs.

Measure halo ellipticity via TMT, \citet{evslin2016}
